Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida

### 2022-01-28 revision/review_response.Rmd

This is an interesting study in that it utilizes a quasi-Lagrangian approach to quantify the influence of a nutrient-rich waste plume on an estuarine ecosystem and is the first that I am aware of that documents the impacts of the Piney Point episode. Nonetheless, there are some points that require clarification to give the reader confidence in the findings. In general, more details are needed on several aspects of the methodology, and I provide specific details below on this as well as other issues.

* __Response__: Thank you for your constructive comments.  Please see the detailed responses below.

Page 1, Line 38 - "Elevated levels of phytoplankton" - needs to be clearer… is this referring to biomass or abundance?

* __Response__: This line was changed to "An initial phytoplankton bloom (non-harmful diatoms) was first observed...".  This was based on chlorohpyll measurements in the vicinity of Port Manataee and cell concentrations enumerated to taxa from laboratory analysis.  The word limit for the abstract prevents these details, but they are presented in text (i.e., line xxx and xxx).

@@

Page 7, Line 22 - Were the breakpoints for K. brevis, or total phytoplankton abundance?

* __Response__: These were applied to K. brevis.  The sentence was revised: "Bloom sizes for HAB species were described qualitatively as low/medium/high concentrations..."

Page 7, Line 37 - What about from ungauged flows?

* __Response__: Our methods for estimating inflows into Tampa Bay are based on methods that are similarly presented in a technical report published by the Tampa Bay Estuary Program in 2012 [@Janicki12].  This report provides the basis for compliance assessment of our TMDL for reporting to the Florida Department of Environmental Protection and represents our best estimate at hydrologic load inputs to the bay.  We have provided a citation to this report in the paper.

@@

Page 11, Line 30 - Need to put Anna Maria Sound on the map. Likewise for Port Manatee.

* __Response__: Locations were added to the map.

Page 12-14, Water quality trends section - Very little attention is given to the role of the hurricane that hit the area in early July. Would be interesting to hear more about its effects on water quality in the study area.

* __Response__: As noted in the red tide section and discussion, Tropical Storm Elsa was a confounding factor when interpreting drivers of the red tide event. It is challenging to disentangle an isolated effect of the storm on the red tide and the same could be said for the broader suite of parameters that were used to assess water quality. For example, a clear chlorophyll spike in early July coincided with the red tide peak, which also coincided with passage of the storm.  A lack of continuous monitoring in the bay also prevented a more comprehensive assessment of potential storm effects.  However, water quality conditions were changing rapidly from the red tide, preceding the arrival of the storm by several days, and we consider the water quality conditions at the time of passage to be driven primarily by the *K. brevis* bloom in the bay. We have added some additional text describing these interpretations in response to your comment below about dissolved oxygen and fish kills.

@@

Page 14, Macroalgae and seagrass trends section - This appears to be one of the weaker components of the project, although it is outside of my expertise. My biggest concern is that it is not clear to me how useful "% occurrence" is. As far as I can tell, this is not equivalent to biomass or areal coverage. Perhaps I am mistaken? Regardless, it would be useful to see a better justification for this metric. Also, how did the 2021 data compare to historic spatial-temporal trends?

* __Response__: Frequency occurrence is a routine reporting metric used by the Tampa Bay Estuary Program to summarize the long-term annual transect monitoring [see @Johansson16;@Sherwood17]. Additional text on line xxx was added to make this clear:

     "Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17]."

     We realize that additional metrics could have been used to summarize the data and we presented trend assessments using the Braun-Blanquet abundance estimates as confirmation for the results from the frequency occurrence analyses (line xxx).  However, this does not provide an assessment of biomass or coverage.  Weight data were collected for select macroalgae samples during the 2021 surveys and a preliminary analysis (see the Seagrass -> Biomass estimates tab [here](https://tbep-tech.github.io/piney-point-analysis/trend)) suggested similar trends as the frequnecy occurrence results.  Because we did not have comparable weight data for seagrasses, the analysis was not included.  It is also worth mentioning that coverage estimates are obtained every two years for Tampa Bay as an annual snapshot (noted on line xxx in the discussion), but these are not of sufficient temporal resolution to support the analyses in this paper.

     There were limitations to our analysis that were noted in the original draft, primarily the lack of long-term historical data on seasonal trends for seagrasses and macroalgae.  For example, line xxx in the original discussion: "Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal and interannual changes and those in potential response to the Piney Point release."

@@

Page 17, Lines 17-20 - How did the 2021 fish kills compare to historical kill spatial-temporal trends?

* __Response__: We were also interested in this question and had included a long-term trend plot in an earlier draft, with results as early as 1995 (see [here](https://github.com/tbep-tech/piney-point-manu/blob/c45049af6885e1e7d29aeb414d161a0ce4bdeadb/figs/redtide.jpeg)).  However, this was not included due to differences in reporting methods that have changed over time.  Online reporting became an option in recent years and we considered the results incomparable to the earlier data that were based on phone reporting methods.

@@

* __Response__: We included some text describing this issue in the discussion.

     "Water quality conditions before and after passage of tropical storm Elsa may have also contributed to fish kills by reducing bottom-water dissolved oxygen. @Stevens06 documented impacts of a category 4 storm on fish resources in the Charlotte Harbor estuary, although tropical storm Elsa was much smaller and fish kills were documented prior to and after arrival of the storm.  Lack of continuous monitoring data for bottom waters in Tampa Bay prevents a more detailed assessment of impacts of the storm on water quality."

Discussion - In general, it would be useful if the team would spend some time establishing the prevalence of N-limitation in the system, either based on their own data or previous studies. The interpretation of plume impacts is predicated on the system being N-limited, so that needs to be established up front.

* __Response__: We agree that the nutrient limitation needs to be clearly stated as the basis for focusing primarily on nitrogen and not phosphorus. The second paragraph in the introduction was meant to provide this justification, but this was clearly insufficient as noted in the review.  We include some additional text and citations in the final paragraph of the introduction as further justification:

@@

     "Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14]. As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production."

     Phosphorus summary plots are also included in the supplement.

@@

Page 20, Line 22 - Ammonium, not ammonia. Or was pH such that ammonia would be the dominant form?

* __Response__: This is in reference to total ammonia nitrogen, as was provided by the water quality sampling.  We have made a note on line xxx in the methods that any mention of ammonia refers to total ammonia nitrogen.

@@

Page 20, Line 39 - Might be worth having a separate paragraph that talks about the timing/duration of the K. brevis bloom in 2021 compared to historic conditions.

* __Response__: Previous content in the discussion below this paragraph provided some historical context on red tide events in the region.  Specifically, starting on line xxx: "Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf..."

     We have added some text below the existing pararaph to provide more context: "Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21]. Severity in estuaries is typically less than Gulf waters as *K. brevis* is limited to higher salinity."

@@

Page 20 (last paragraph)-21 (first paragraph) - Without showing methods and/or data, I feel that it is inappropriate to introduce the nekton abundance/composition here.

* __Response__: Agreed, much of this content was removed.

Page 21, Lines 32-47 - Much of this text on seagrass seems speculative, and again, unclear about use of % occurrence as a metric.

* __Response__: We have replaced the beginning of this paragraph with more definitive text:

     "Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein]. The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses."

     Also, please see our response above about frequency occurrence.

### 2022-01-28 manu-draft.Rmd
rev4 comments addressed

Water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).  Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release. Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP and those of the inter-agency partners.  Data quality objectives followed guidelines outlined in the USEPA-approved TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of split samples between laboratories to check precision of measured values. Samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, Secchi depth, etc.).  In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts, depending on monitoring agency.  Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.

@@

Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For water quality variables, monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.  Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist. Methods used to accommodate measured concentrations of water quality variables that were below detection included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].

@@

For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were not observed over the initial study period. The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern [e.g., alteration of sediment geochemistry, @Eldridge04]. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa* sp. has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges to managing seagrass health.

@@

Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K. brevis* bloom observed in Tampa Bay this year is difficult in the absence of more quantitative results or mechanistic tools to support understanding.  Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay. Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger71], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].  Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, also created conditions that were favorable for growth of *K. brevis*. However, the results suggest a likely scenario that residual nutrients from the Piney Point release, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K. brevis* to the concentrations observed in July [also see @Medina20].  Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the plume was widespread throughout the bay and persisted for many months after the release ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production. Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.

Additional monitoring and analysis are also required to fully understand the long-term impacts to bay resources beyond water quality.  For benthic communities, sediments sampled in April and September near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106, see https://shiny.tbep.org/piney-point], possibly reflecting the generally high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].  Comparison of the April and September samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([TBEP unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)). Differences between the April and September samples were not observed. Finally, effects of changing environmental conditions and *K. brevis* on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources.  Twenty manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021. This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that are likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.

>>>>

Water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L, hereafter referred to as ammonia), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).  Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release. Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP and those of the inter-agency partners.  Data quality objectives followed guidelines outlined in the USEPA-approved TBEP Data Quality Management Plan [@tbep1620]. Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of split samples between laboratories to check precision of measured values. Samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, Secchi depth, etc.).  In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts, depending on monitoring agency.  Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.

@@

Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17].  Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For water quality variables, monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.  Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist. Methods used to accommodate measured concentrations of water quality variables that were below detection included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].

@@

Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein]. The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses.  The long-term effects of the Piney Point discharge on the seagrass community remains uncertain. From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).  These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats. Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K. brevis* bloom in 2018. Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern [e.g., alteration of sediment geochemistry, @Eldridge04]. Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from annual transect monitoring results for Tampa Bay.  In particular, increasing abundance in recent years of the green algae *Caulerpa* sp. has been observed at long-term transects that were previously dominated by seagrass.  These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06]. Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges to managing seagrass health.

@@

Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K. brevis* bloom observed in Tampa Bay this year is difficult in the absence of more quantitative results or mechanistic tools to support understanding.  Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay. Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger71], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].  Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21]. Severity in estuaries is typically less than Gulf waters as *K. brevis* is limited to higher salinity.  Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, created conditions that were favorable for growth of *K. brevis* in Tampa Bay. However, the results suggest a likely scenario that residual nutrients from the Piney Point release, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K. brevis* to the concentrations observed in July [also see @Medina20].  Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the plume was widespread throughout the bay and persisted for many months after the release ceased at Port Manatee.  Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production. Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.
